### 2023-10-05 proposal/proposal.tex
added sections/outline to proposal

\documentclass[12pt]{article}

%% preamble: Keep it clean; only include those you need
\usepackage{amsmath}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}

% for space filling
\usepackage{lipsum}
% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}

%% meta data

\title{Proposal: }
\author{Luke Noel\\
  Department of Statistics\\
  University of Connecticut
}

\begin{document}
\maketitle

\paragraph{Introduction}

\paragraph{Specific Aims}

\paragraph{Data}

\paragraph{Research Design and Methods}

\paragraph{Discussion}

\bibliography{../manuscript/refs}
\bibliographystyle{chicago}

\end{document}

### 2023-10-10 proposal/proposal.tex
finished intro

@@
 
-%% preamble: Keep it clean; only include those you need
+%% preamble:
 \usepackage{amsmath}
@@
 
-% for space filling
-\usepackage{lipsum}
 % highlighting hyper links
@@
 
-\title{Proposal: }
+\title{Proposal: Evaluating the Impact of 2023 MLB Rule Changes on Performance
+ Metrics and Strategies in Major League Baseball}
 \author{Luke Noel\\
@@
 \paragraph{Introduction}
+Major League Baseball (MLB) is a dynamic sport that continually evolves through 
+rule changes aimed at enhancing game pace, player safety, and fan engagement. In
+the 2023 MLB season, significant rule changes were introduced, including the 
+implementation of a pitch clock, outlawing the defensive shift, and larger 
+bases. This research proposal outlines a comprehensive study that aims to
+analyze the impact of these rule changes on various performance metrics in
+baseball, including but not limited to stolen bases, pace of play,
+and batting averages.
 

### 2023-10-10 proposal/proposal.tex
finished aims, data, and methods sections of proposal

@@
 the 2023 MLB season, significant rule changes were introduced, including the 
-implementation of a pitch clock, outlawing the defensive shift, and larger 
+implementation of a pitch clock, the ban of the defensive shift, and larger 
 bases. This research proposal outlines a comprehensive study that aims to
@@
 \paragraph{Specific Aims}
+The specific aims of this paper are to investigate the impact of the 2023 MLB
+rule changes on various performance metrics and to assess whether these changes
+had significant effects. Some research questions I will address are:
+
+\begin{enumerate}
+  \item How did the 2023 MLB rule changes influence the frequency and success 
+  rate of stolen base attempts?
+  \item Were there statistically significant changes in time of play before and
+   after the rule changes?
+  \item Did the rule changes have a noticeable impact on batting averages and 
+  other offensive metrics?
+  \item What are the effect sizes associated with the observed differences
+   in these metrics?
+\end{enumerate}
 
 \paragraph{Data}
+The data for this research will consist of comprehensive statistics for the 2023
+MLB season, including game logs, and player/team/league performance metrics.
+This data will then be compared and contrasted with the same data but 
+from previous seasons where these rule changes were not implemented yet. There
+are various sources where the data will be obtained from: \href{https://www.espn.com/mlb/stats/team/_/season/2023/seasontype/2}{ESPN}
+and \href{https://www.baseball-reference.com/}{Baseball Reference} will give
+league, team, and player averages across seasons along with game duration.
+\href{https://baseballsavant.mlb.com/league?season=2022#statcastHitting}{Baseball Savant}
+will be useful for providing advanced statistics.
 
 \paragraph{Research Design and Methods}
+To address the research questions, a combination of statistical methods will be
+used. This may include two sample hypothesis tests, regression analysis, and
+various data visualization techniques to truly show the impact of the rule
+changes. Each performance metric such as batting average, home runs, and stolen
+bases will be analyzed separately to determine whether the rule changes had a
+statistically significant impact.
 

### 2023-10-10 proposal/proposal.tex
finished discussion section of proposal

@@
 \paragraph{Discussion}
+The most challenging aspect of this research project will be accounting for 
+confounding factors such as changes in player rosters which might influence the
+observed differences in the metrics. This goes hand in hand with the limitations
+which include the assumption that the observed differences are solely attributable
+to the rule changes, without considering other external factors. Another limitation
+is the amount of work done on this topic, as the rule changes were just implemented
+in 2023. There are some articles that I can reference such as ..ref.. and 
+..ref.. that talk about the effects of the rule changes, but no scholarly papers
+have been posted about it yet. I will have to research more about statistical
+papers dedicated to analyzing changes with MLB stats between seasons, in general.
+
+If challenges arise, a fall back plan will involve conducting sensitivity analyses
+and exploring alternative statistical tests/visualizations. All in all, I believe
+this paper will be a great resource to show the impact of the rule changes on 
+the game of baseball.
 

### 2023-10-10 proposal/proposal.tex
added citations to proposal

@@
 is the amount of work done on this topic, as the rule changes were just implemented
-in 2023. There are some articles that I can reference such as ..ref.. and 
-..ref.. that talk about the effects of the rule changes, but no scholarly papers
-have been posted about it yet. I will have to research more about statistical
-papers dedicated to analyzing changes with MLB stats between seasons, in general.
+in 2023. There are some articles that I can reference such as \citet{Connon} and
+\citet{Castrovince} that talk about the effects of the rule changes, but no 
+scholarly papershave been posted about it yet. I will have to research more about
+statistical papers dedicated to analyzing changes with MLB stats between seasons, 
+in general.
 

### 2023-10-25 manuscript/paper.tex
added metadata to manuscript

--- 2023-10-10 proposal/proposal.tex
@@
 
-%% preamble:
+%% preamble: Keep it clean; only include those you need
 \usepackage{amsmath}
@@
 
-\title{Proposal: Evaluating the Impact of 2023 MLB Rule Changes on Performance
- Metrics and Strategies in Major League Baseball}
+\title{Evaluating the Impact of 2023 Rule Changes in Major League Baseball}
 \author{Luke Noel\\
@@
 }
-
-\begin{document}
-\maketitle
-
-\paragraph{Introduction}
-Major League Baseball (MLB) is a dynamic sport that continually evolves through 
-rule changes aimed at enhancing game pace, player safety, and fan engagement. In
-the 2023 MLB season, significant rule changes were introduced, including the 
-implementation of a pitch clock, the ban of the defensive shift, and larger 
-bases. This research proposal outlines a comprehensive study that aims to
-analyze the impact of these rule changes on various performance metrics in
-baseball, including but not limited to stolen bases, pace of play,
-and batting averages.
-
-\paragraph{Specific Aims}
-The specific aims of this paper are to investigate the impact of the 2023 MLB
-rule changes on various performance metrics and to assess whether these changes
-had significant effects. Some research questions I will address are:
-
-\begin{enumerate}
-  \item How did the 2023 MLB rule changes influence the frequency and success 
-  rate of stolen base attempts?
-  \item Were there statistically significant changes in time of play before and
-   after the rule changes?
-  \item Did the rule changes have a noticeable impact on batting averages and 
-  other offensive metrics?
-  \item What are the effect sizes associated with the observed differences
-   in these metrics?
-\end{enumerate}
-
-\paragraph{Data}
-The data for this research will consist of comprehensive statistics for the 2023
-MLB season, including game logs, and player/team/league performance metrics.
-This data will then be compared and contrasted with the same data but 
-from previous seasons where these rule changes were not implemented yet. There
-are various sources where the data will be obtained from: \href{https://www.espn.com/mlb/stats/team/_/season/2023/seasontype/2}{ESPN}
-and \href{https://www.baseball-reference.com/}{Baseball Reference} will give
-league, team, and player averages across seasons along with game duration.
-\href{https://baseballsavant.mlb.com/league?season=2022#statcastHitting}{Baseball Savant}
-will be useful for providing advanced statistics.
-
-\paragraph{Research Design and Methods}
-To address the research questions, a combination of statistical methods will be
-used. This may include two sample hypothesis tests, regression analysis, and
-various data visualization techniques to truly show the impact of the rule
-changes. Each performance metric such as batting average, home runs, and stolen
-bases will be analyzed separately to determine whether the rule changes had a
-statistically significant impact.
-
-\paragraph{Discussion}
-
-\bibliography{../manuscript/refs}
-\bibliographystyle{chicago}
-
-\end{document}

### 2023-10-26 manuscript/paper.tex
added abstract section and title
modified the title, started the abstract

@@
 
-\title{Evaluating the Impact of 2023 Rule Changes in Major League Baseball}
+\title{The Effects of the 2023 Rule Changes in Major League Baseball}
 \author{Luke Noel\\
@@
 }
+
+\begin{document}
+\maketitle
+
+\begin{abstract}
+Major League Baseball (MLB) is a sport continuously evolving through rule
+changes dedicated to improving game dynamics, player safety, and fan engagement.
+Prior to the start of the 2023 season, MLB's commissioner Rob Manfred
+implemented a set of significant rule changes, headlined by the introduction of 
+a pitch clock, defensive shift restrictions, and bigger bases.
+
+\end{abstract}
+
+\end{document}

### 2023-10-26 manuscript/paper.tex
finished abstract

@@
 implemented a set of significant rule changes, headlined by the introduction of 
-a pitch clock, defensive shift restrictions, and bigger bases.
-
+a pitch clock, defensive shift restrictions, and bigger bases. Although these
+changes were intended to improve pace of play, action, and safety of the game,
+their actual impact on various aspects of baseball performance and strategy is 
+not well understood. This study uses statistical methods to evaluate the effects
+of the rule changes on game duration, scoring, hitting, pitching, fielding, and
+baserunning. We compare the 2023 season data with the previous seasons using
+descriptive statistics, hypothesis testing, and regression analysis. INSERT
+major findings here.
 \end{abstract}

### 2023-10-26 manuscript/paper.tex
started intro, added source to refs

@@
 
+\section{Introduction}
+\label{sec:intro}
+
+Major League Baseball (MLB), often referred to as America's pastime, is a sport
+rich in tradition and innovation. Throughout its long history, MLB has enacted
+numerous rule changes aimed at improving the game's relevance, competitiveness,
+and entertainment value. For instance, in 1973, the designated hitter rule was
+implemented for all American League teams, with the hopes of increasing
+offensive production as a whole. However, we are not really interested in "why"
+the rules are changed, but more the underlying effects of the change(s). This
+can be seen in \citet{Cooley} where the effects of the designated hitter rule
+on player longevity, pitcher complete games, and number of hit batters are
+analyzed.
+
 \end{document}

### 2023-10-26 manuscript/paper.tex
added items to intro

@@
 on player longevity, pitcher complete games, and number of hit batters are
-analyzed.
+analyzed. We see many more major and minor MLB rule changes come into action
+throughout the course of its history, like the recent "ghost runner" rule carried
+out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
+automatically starts out on second-base in extra-innings.
+
+Building upon this willingness for change, prior to the start of the 2023 season,
+Rob Manfred, MLB's commissioner, implemented a set of significant rule changes
+intended to improve the pace of play, action, and safety of the game. Here is
+a list of the new rules:
+
+\begin{itemize}
+  \item \textbf{Pitch Timer:} 30-second timer between batters. Between pitches,
+  a 15-second timer with the bases empty and a 20-second timer with runners on base.
+  Pitchers who violate the timer are charged with an automatic ball.
+  Batters who violate the timer are charged with an automatic strike.
+
+  \item \textbf{Defensive Shift Limits:} The defensive team must have a minimum
+  of four players on the infield, with at least two infielders completely on either
+  side of second base.
+
+  \item \textbf{Bigger Bases:} The bases, which traditionally have been 15 inches
+  square, will instead be 18 inches square.
+
+  \item \textbf{Pitcher Disengagement Limits:} Pitchers are limited to two
+  disengagements (pickoff attempts or step-offs) per plate appearance. If a third
+  pickoff attempt is made, the runner automatically advances one base if the 
+  pickoff attempt is not successful.
+\end{itemize}
 

### 2023-10-30 manuscript/paper.tex
added/updated a reference

@@
 implemented for all American League teams, with the hopes of increasing
-offensive production as a whole. However, we are not really interested in "why"
+offensive production as a whole. However, we are not really interested in \emph{why}
 the rules are changed, but more the underlying effects of the change(s). This
@@
 analyzed. We see many more major and minor MLB rule changes come into action
-throughout the course of its history, like the recent "ghost runner" rule carried
-out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
+throughout the course of its history, like the recent \emph{ghost runner} rule
+carried out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
 automatically starts out on second-base in extra-innings.
@@
 intended to improve the pace of play, action, and safety of the game. Here is
-a list of the new rules:
+a list of the new rules and a short description for each, according to \citet{Castrovince}:
 
@@
 
+\bibliographystyle{chicago}
+\bibliography{refs.bib}
+
 \end{document}

### 2023-10-30 manuscript/paper.tex
finished intro

@@
 
+These rule changes may seem ridiculous to many baseball \emph{purists} out there,
+but here is a quote from Commissioner Manfred that explains the rationale:
+
+\begin{quote}
+  We've tried to address the concerns expressed in a thoughtful way, respectful,
+  always, of the history and traditions of the game, and of player concerns. Our
+  guiding star in thinking about changes to the game has always been our fans.
+  What do our fans want to see on the field? We've conducted thorough and ongoing
+  research with our fans, and certain things are really clear. Number 1, fans want
+  games with better pace. Two, fans want more action, more balls in play. And three,
+  fans want to see more of the athleticism of our great players.
+\end{quote}
+
+Also, keep in mind that these rule changes have been thoroughly tested and refined
+in the Minor Leagues for many years so it's not like it was a split decision to
+implement them in the Majors. Each of the rules were actually tested in about 8,000
+Minor League games \citep{Castrovince}.
+
+The next section is a Literature Review that details some relevant works on the
+topic, but it is important to know that the data is so new that the impact of the
+rule changes on various aspects of baseball performance metrics and strategies 
+is not well understood yet. This study uses statistical methods to evaluate the
+effects of the rule changes on game duration, scoring, hitting, pitching, fielding,
+and baserunning. For instance, a major baseball statistic that is analyzed is
+stolen bases (SB) which we should see an increase of across the league due to the
+rule changes. Bigger bases mean slightly less distance traveled to get the
+SB, and the pitch timer means that baserunners can time up the pitchers easier
+and know when to start stealing. We compare the 2023 season data with the previous
+seasons using descriptive statistics, hypothesis testing, and regression analysis.
+
+This analysis is important as it holds the potential to inform future rule changes,
+offering empirical insights that can guide decision-makers in shaping the game's
+future. The data and findings generated by this research may serve as a valuable
+resource for MLB executives, team managers, and players in adapting to the new
+rule landscape. Equally crucial, fans and enthusiasts will gain deeper insights
+into the impact of these changes on the sport they love, providing context to
+their experiences as spectators.
+
+% roadmap
+The rest of the paper is organized as follows.
+A literature review of relevant works will be examined in Section~\ref{sec:lit}.
+The data used in the study will be presented in Section~\ref{sec:data}.
+The methods are described in Section~\ref{sec:meth}.
+The results are reported in Section~\ref{sec:resu}.
+A discussion concludes in Section~\ref{sec:disc}.
+
 \bibliographystyle{chicago}

### 2023-10-30 manuscript/paper.tex
added outline for rest of paper

@@
 
+\section{Literature Review}
+\label{sec:lit}
+
+Examine relevant works.
+
+\section{Data}
+\label{sec:data}
+
+Use this section to describe the data that helps to answer your research
+questions.
+
+\section{Methods}
+\label{sec:meth}
+
+Use this section to present the methodologies that will generate results by
+analyzing the data.
+
+\section{Results}
+\label{sec:resu}
+
+Summarize the results.
+
+\section{Discussion}
+\label{sec:disc}
+
+What are the main contributions again?
+
+What are the limitations of this study?
+
+What are worth pursuing further in the future?
+
 \bibliographystyle{chicago}

### 2023-11-07 manuscript/paper.tex
finished literature review section, added some refs to bib

@@
 
-Examine relevant works.
+As stated previously, there has been no scholarly literature yet on the effects
+of the 2023 MLB rule changes (as of fall 2023), since it is so new. However,
+there has been research done already on how a certain rule change/changes
+affected a sport. For example, as referenced above, \citet{Cooley} investigated
+the impact of the Designated Hitter (DH) in the MLB, a rule change that let a batter
+hit for a pitcher in the lineup, instituted in 1973. They analyzed how the
+implementation of the DH impacted certain metrics like attendance, complete games
+by pitchers (CG), and total runs scored. They did so using various statistical
+methods such as correlation analysis, regression (particularly r-squared analysis),
+and compound average growth rates. They concluded that the DH led to more offensive
+production, a decrease in total complete games by pitchers, and had no significant
+effect on overall fan attendance at games.
+
+However, we do not need to just look at Major League Baseball to see
+the effects of some rule changes in sports. For instance, \citet{Madera}
+discusses the impact of certain rule changes on game-related statistics in
+professional men's water polo matches. They did so by calculating means and
+standard deviations for certain game-related stats such as possession time,
+total goals, etc., and running one-way ANOVA to test for significance in the
+difference of these metrics pre and post rule changes. In a similar fashion,
+\citet{Nourayi} analyzed the effects of the new defensive 3-second rule and
+elimination of hand checking in the National Basketball Association (NBA), rule
+changes aimed at improving the flow and pace of the game. This was done using
+many different t-tests which resulted in finding that the total FGA (Field Goals
+Attempted) and 3FGA (3-Point Field Goals Attempted) were significantly greater
+in the post rule change period.
+
+Also, we do not need to just look at rule changes specifically, we can explore
+a \emph{before and after analysis} of some other change made in sports. \citet{Price}
+is a good illustration of this is, as they conducted an analysis of the COVID
+\emph{bubble} in the NBA playoffs and how this affected home-court advantage.
+Nine different z-tests were conducted to try to see if there was significance in
+the difference in free-throw rate, winning percentage, etc. in the 2020 season
+compared to previous years. This highly relates to our study, but we will instead
+be testing the difference in say stolen bases for example, from the 2023 season
+compared to previous years. \citet{Zhang} is another great example as they
+focused on the effect of the video assistant referee (VAR) on referee's decisions
+in the FIFA Women's World Cup. They used a Mann-Whitney U test to compare differences
+in game-related variables such as playing time, penalties, and goals, with and
+without VAR.
 

### 2023-11-09 manuscript/paper.tex
added a couple paragraphs to data section

@@
 implemented for all American League teams, with the hopes of increasing
-offensive production as a whole. However, we are not really interested in \emph{why}
+offensive production as a whole. However, we are not really interested in ``why''
 the rules are changed, but more the underlying effects of the change(s). This
@@
 analyzed. We see many more major and minor MLB rule changes come into action
-throughout the course of its history, like the recent \emph{ghost runner} rule
+throughout the course of its history, like the recent ``ghost runner'' rule
 carried out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
@@
 
-These rule changes may seem ridiculous to many baseball \emph{purists} out there,
+These rule changes may seem ridiculous to many baseball ``purists'' out there,
 but here is a quote from Commissioner Manfred that explains the rationale:
@@
 Also, we do not need to just look at rule changes specifically, we can explore
-a \emph{before and after analysis} of some other change made in sports. \citet{Price}
+a ``before and after analysis'' of some other change made in sports. \citet{Price}
 is a good illustration of this is, as they conducted an analysis of the COVID
-\emph{bubble} in the NBA playoffs and how this affected home-court advantage.
+``bubble'' in the NBA playoffs and how this affected home-court advantage.
 Nine different z-tests were conducted to try to see if there was significance in
@@
 
-Use this section to describe the data that helps to answer your research
-questions.
+The data used for this analysis was collected from \href{https://baseballsavant.mlb.com/}{Baseball Savant},
+a comprehensive and interactive platform that contains advanced statistics and
+data related to Major League Baseball (MLB). The main variables of interest
+are Stolen Bases (SB), Batting average (BA), Strike Outs (SO), and Batting
+Average on Balls in Play (BABIP), among some others. These are game-related
+baseball statistics that will be analyzed similar to those used in the literature
+above, like we saw with Free Throw Rate in the NBA, for example. These variables
+were selected based on domain knowledge and intuition about what metrics may be
+affected most by the new MLB rule changes. We expect to see Stolen Bases increase
+across the league, due to the pitch clock and pitcher disengagement limits. Batting
+Average, Batting Average on Balls in Play, and Slugging (SLG) may be affected by
+the limit of the defensive shift, as this rule change may increase the likelihood
+of a hit. In a similar vein, Strike Outs and Walks (BB) may see an impact as both
+pitchers and batters adjust to the pitch clock. Other variables included in the
+study:
+\begin{itemize}
+  \item Ground Into Double Play (GIDP)
+  \item Hits on Ground Balls (Hit-GB)
+  \item Expected Batting Average (xBA)
+\end{itemize}
+We expect these variables will be affected by the limit of the defensive shift.
+Also, pace of play (game time) data was collected from \href{https://www.baseball-reference.com/}{Baseball Reference}.
+Each row of this dataset represents a single season from 2023 dating back to 1998
+(26 observations total), containing the Average Game Time (in hours) for that season.
+Intuitively, we should expect this variable to have a significant decrease for the
+2023 season, due to the introduction of the pitch clock making the games go faster.
+Ideally, we would have used datasets containing the game times for \underline{each}
+game in both seasons, however this was not publicly available.
+
+The main data used though was collected using the Custom Leaderboard feature
+on Baseball Savant, where each variable above was added to the dataset, with each
+row representing a single player's overall statistics. The data was then filtered
+to only include the 2023 season, and only include players with a minimum 100 Plate Appearances (PA)
+that season to limit outliers (100 min. PA is a popular threshold used for 
+baseball player analysis). This gives us 461 observations for the 2023 season,
+again with each row being a unique batter. The same steps were then applied to
+create a comparison dataset now filtering for the 2022 season, which has 469
+observations total. This gives us a baseline to compare the variables in the
+season where the rule changes were in effect (2023), to the previous season where
+they were not (2022).
+
+Different data formats were considered, such as using team-wide statistics
+for each season instead of player-wide, where the dataset would be 30 rows (one
+for each team). Using a combination of previous seasons data for comparison (not
+just 2022) was considered also, but we settled on the data we used as it seemed
+the most straightforward for our analysis and limited potential confounding
+factors. Data from complete game logs for each game of the 2023 season would have
+been an option, however this data is not yet publicly available. More on these topics
+are discussed in Section's~\ref{sec:meth} and \ref{sec:disc}.
 

### 2023-11-09 manuscript/paper.tex
added data folder with 3 datasets

@@
 study:
+
 \begin{itemize}
@@
 \end{itemize}
+
 We expect these variables will be affected by the limit of the defensive shift.
@@
 season where the rule changes were in effect (2023), to the previous season where
-they were not (2022).
+they were not (2022). There are no missing or null values in each dataset.
 
@@
 factors. Data from complete game logs for each game of the 2023 season would have
-been an option, however this data is not yet publicly available. More on these topics
-are discussed in Section's~\ref{sec:meth} and \ref{sec:disc}.
+been an option, however this data is not yet publicly available. More on this topic
+is discussed in Section~\ref{sec:disc}.
 

### 2023-11-09 code/barplot_glength.R
added code folder with a script for a barplot

@@ -1,0 +1,27 @@
+library(tidyverse)
+library(ggplot2)
+library(viridis)
+
+g_times <- read.csv("../data/game_times.csv")
+
+# Converting AvgTime into a decimal
+g_times <- g_times %>%
+  separate(AvgTime, into = c("hours", "minutes"), sep = ":") %>%
+  mutate(
+    hours = as.numeric(hours),
+    minutes = as.numeric(minutes)
+  )
+g_times$decimal_time <- g_times$hours + g_times$minutes / 60
+
+# Creating barplot
+g_times$Year <- as.factor(g_times$Year)
+
+g_length <- ggplot(g_times, aes(x = Year, y = decimal_time, fill = Year)) +
+  geom_bar(stat = "identity", position = "identity", color = "white", size = 1.5) +
+  scale_fill_viridis(discrete = TRUE, option = "viridis") +
+  labs(x = "Year",
+       y = "Average Length of Game (Hours)") +
+  theme_minimal() +
+  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggsave("barplot_glength.pdf", g_length, height=4, width=6)

### 2023-11-09 code/barplot_glength.R
updated paths in barplot script

@@ -2,7 +2,7 @@ library(tidyverse)
 library(ggplot2)
 library(viridis)
 
-g_times <- read.csv("../data/game_times.csv")
+g_times <- read.csv("course-paper-or-presentation-lukenoel20/data/game_times.csv")
 
 # Converting AvgTime into a decimal
 g_times <- g_times %>%
@@ -17,11 +17,11 @@ g_times$decimal_time <- g_times$hours + g_times$minutes / 60
 g_times$Year <- as.factor(g_times$Year)
 
 g_length <- ggplot(g_times, aes(x = Year, y = decimal_time, fill = Year)) +
-  geom_bar(stat = "identity", position = "identity", color = "white", size = 1.5) +
+  geom_bar(stat = "identity", position = "identity", color = "white", linewidth = 1.5) +
   scale_fill_viridis(discrete = TRUE, option = "viridis") +
   labs(x = "Year",
        y = "Average Length of Game (Hours)") +
   theme_minimal() +
   theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
 
-ggsave("barplot_glength.pdf", g_length, height=4, width=6)
+ggsave("barplot_glength.pdf", g_length, path="course-paper-or-presentation-lukenoel20/manuscript", height=4.2, width=6)

### 2023-11-09 manuscript/paper.tex
added barplot figure and description to manuscript data section

@@
 
+One of the main reasons commissioner Rob Manfred enacted the pitch clock ``rule''
+was to speed up the games and improve the pace of play to make them more engaging
+for the fans, and preliminary analysis shows this worked. Notice in Figure~\ref{fig:Fig1}
+the average length of games for the 2023 season was the shortest in the past 25 years.
+There is some season-to-season variation but the highlighted yellow bar on the right
+(representing the 2023 season) seems significantly lower than the rest. We will
+test if the difference in average game length between 2023 and 2022 is statistically
+significant in Section~\ref{sec:meth}.
+
+\begin{figure}[tbp]
+  \centering
+  \includegraphics{barplot_glength}
+  \caption{Average Game Length in hours for all MLB games from 1998 to 2023.}
+  \label{fig:Fig1}
+\end{figure}
+
 \section{Methods}

### 2023-11-10 code/sb_hist.R
added histogram script to code folder

@@ -1,0 +1,22 @@
+library(ggplot2)
+
+p_2023 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2023.csv")
+p_2022 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2022.csv")
+
+# Combine both datasets for comparison
+df_combined <- rbind(cbind(p_2023, Season = "2023"), cbind(p_2022, Season = "2022"))
+
+# Convert the SB column to numeric
+df_combined$r_total_stolen_base <- as.numeric(df_combined$r_total_stolen_base)
+
+# Creating overlayed histograms for stolen bases (2022 vs. 2023)
+sb_hist <- ggplot(df_combined, aes(x = r_total_stolen_base, fill = factor(year), color = factor(year))) +
+  geom_histogram(position = "dodge", bins = 30, alpha = 0.7, stat = "count") +
+  labs(
+       x = "Stolen Bases",
+       y = "Count") +
+  scale_fill_manual(name = "Year", values = c("2023" = "blue", "2022" = "orange")) +
+  scale_color_manual(name = "Year", values = c("2023" = "darkblue", "2022" = "darkorange")) +
+  theme_minimal()
+
+ggsave("sb_hist.pdf", sb_hist, path = "course-paper-or-presentation-lukenoel20/manuscript", height=4.2, width=6)

### 2023-11-10 manuscript/paper.tex
added histogram figure and description to manuscript

@@
 
+To visually compare 2023 baseball statistics versus those in 2022, two histograms
+were created for each variable with 2023's data overlaid on 2022's. Figure~\ref{fig:Fig2}
+illustrates an example of this for the Stolen Bases variable, as we think this
+metric may be most affected by the rule changes, as stated previously. Both
+histograms exhibit fairly similar distributions; however, in 2023, there is a
+noticeable increase in the number of players with 20 or more stolen bases. This
+observation supports our intuition that the rule changes contributed to an uptick
+in this variable.
+
+\begin{figure}[tbp]
+  \centering
+  \includegraphics{sb_hist}
+  \caption{Comparison of Stolen Bases between the 2023 and 2022 MLB Seasons. The
+  histograms illustrate the distribution of stolen base counts for players in each
+  season, with blue representing 2023 and orange representing 2022.}
+  \label{fig:Fig2}
+\end{figure}
+
 \section{Methods}

### 2023-11-10 manuscript/paper.tex
added boxplots code to code folder, included it in manuscript with description

@@
 \usepackage{natbib}
+\usepackage{setspace}
 
@@
 \usepackage[colorlinks=true, citecolor=blue]{hyperref}
+
+\doublespacing
 
@@
 observation supports our intuition that the rule changes contributed to an uptick
-in this variable.
+in this variable. Side by side boxplots for 2022 versus 2023 were also created for
+each variable as shown in Figure~\ref{fig:Fig3}. The Stolen Bases statistic again
+sticks out, having a greater median and more high outliers in 2023. Expected Batting Average (XBA)
+and Slugging Percentage (SLG Percent) both look intriguing as well. As said above
+with Game Length, the statistical significance of the difference in these statistics
+from 2022 to 2023 will be tested in Section~\ref{sec:meth}.
 
@@
   \label{fig:Fig2}
+\end{figure}
+
+\begin{figure}[tbp]
+  \centering
+  \includegraphics{boxplots}
+  \caption{Matrix of boxplots illustrating the evolution of baseball statistics
+  from 2022 to 2023, encompassing strikeout rates, walk rates, batting averages,
+  slugging percentages, BABIP, total stolen bases, GIDP, ground ball hits, and xBA.}
+  \label{fig:Fig3}
 \end{figure}

### 2023-11-10 code/boxplots.R

@@ -1,0 +1,93 @@
+library(ggplot2)
+library(gridExtra)
+
+p_2023 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2023.csv")
+p_2022 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2022.csv")
+
+# Combine both datasets for comparison
+df_combined <- rbind(cbind(p_2023, Season = "2023"), cbind(p_2022, Season = "2022"))
+
+df_combined$year <- as.factor(df_combined$year)
+
+# Create boxplots for each variable 
+plot_strikeout <- ggplot(df_combined, aes(x=year, y=strikeout, fill=year)) +
+  geom_boxplot() +
+  ggtitle("Strikeouts") +
+  theme(legend.position = "none", 
+        axis.title.x = element_blank(), 
+        axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5)) 
+
+plot_walk <- ggplot(df_combined, aes(x=year, y=walk, fill=year)) +
+  geom_boxplot() +
+  ggtitle("Walks") +
+  theme(legend.position = "none", 
+        axis.title.x = element_blank(), 
+        axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5)) 
+
+p_ba <- ggplot(df_combined, aes(x=year, y=batting_avg, fill=year)) +
+  geom_boxplot() +
+  ggtitle("Batting Avg") +
+  theme(legend.position = "none", 
+        axis.title.x = element_blank(), 
+        axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5)) 
+
+p_slg <- ggplot(df_combined, aes(x=year, y=slg_percent, fill=year)) +
+  geom_boxplot() +
+  ggtitle("SLG Percent") +
+  theme(legend.position = "none", 
+        axis.title.x = element_blank(), 
+        axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5)) 
+
+p_babip <- ggplot(df_combined, aes(x=year, y=babip, fill=year)) +
+  geom_boxplot() +
+  ggtitle("BABIP") +
+  theme(legend.position = "none", 
+        axis.title.x = element_blank(), 
+        axis.title.y = element_blank(),
+        plot.title = element_text(hjust = 0.5)) 
+
[…]

### 2023-11-10 manuscript/paper.tex
finished methods section of manuscript

@@
 
-Use this section to present the methodologies that will generate results by
-analyzing the data.
+The introduction of these rule changes in the MLB provides a new and exciting
+opportunity to study how they impacted various aspects of the game. This study
+mainly focuses on comparing baseball statistics from the 2023 season versus the
+2022 season to limit potential confounding factors. If we instead compared the
+post-rule change season to say the last 10 MLB seasons, observed differences
+may not be solely due to the 2023 rule changes, but rather from the effects of
+drastic changes in the game between now and then. For instance, there were other
+rule changes made in the last 10 years, such as the National League (NL) adopting
+the designated hitter in 2022, relief pitchers having to pitch to at least 3 batters,
+etc., that impacted statistics around the league. The 2020 season was also
+drastically different due to COVID-19, with less games being played and no fans in
+the stands. The ``juiced balls'' conspiracy was also running rampant in the last
+decade, with evidence to show that in some seasons, players were playing with
+baseballs that traveled further and harder off the bat. Offensive and defensive
+philosophy has also seen historical changes, with defensive shifting becoming
+more and more prevalent and teams diving more and more into the ``home run or
+bust''/launch angle theory with batting. However, 2022 and 2023 were fairly similar
+in nature all things considered with the obvious change being the new rules implemented,
+which we will test to see how they affected statistics in 2023.
+
+Comparisons between the 2023 and 2022 seasons were made on offensive metrics such
+as batting average, stolen bases, and slugging percentage, along with game length.
+Comparing the difference in these metrics between seasons will reveal valuable insights
+into understanding how the rule changes impacted the game.
+
+The ten following specific research questions will test the effects of the 2023
+MLB rule changes on the game of baseball:
+
+\begin{enumerate}
+  \def\labelenumi{\arabic{enumi}.}
+  \item
+    Is the average game length shorter in 2023 compared to the previous 25 years?
+  \item
+    Are players stealing more bases in 2023 than they did in 2022?
+  \item
+    Is the strikeout rate different in 2023 compared to 2022?
+  \item
+    Is the walk rate different in 2023 compared to 2022?
+  \item
+    Is the batting average different in 2023 compared to 2022?
+  \item
+    Is the slugging percentage different in 2023 compared to 2022?
+  \item
+    Is the batting average on balls in play different in 2023 compared to 2022?
+  \item 
+    Is the expected batting average different in 2023 compared to 2022?
+  \item
+    Is the number of hits on ground balls greater in 2023 compared to 2022?
+  \item
+    Is the number of ground balls into double plays different in 2023 compared to 2022?
+\end{enumerate}
+
+All ten questions can be tested using a standard two-sample comparison with the \(t\)-test.
+The \(z\)-test was considered as done in \citet{Price}, but we do not know the population standard deviation
+in this case. The \(t\)-test assumes the data is normally distributed so to account
+for this, we also conducted the nonparametric, distribution free, Mann Whitney U
+Test (also known as the Wilcoxon Rank Sum Test) to be sure of our results. Also,
+notice how some of these research questions call for a one-sided hypothesis test,
+like the number of hits on ground balls, because we intuitively expect this number
+to be greater in 2023 than in 2022 due to the limit of the defensive shift. Other
+metrics that we were not sure about whether they would increase or decrease were
+tested using a two-sided hypothesis test, simply testing for a significance in the
+overall difference from season to season. Other methods such as regression were
+considered for this study as well, but we are primarily interested in comparing
+means and assessing the effect of a single variable, so a \(t\)-test is more appropriate
+and straightforward.
 

### 2023-11-10 manuscript/paper.tex
added code for table in code folder, added the table to manuscript also

@@
 
+\setlength{\tabcolsep}{10pt}
 \doublespacing
@@
 for this, we also conducted the nonparametric, distribution free, Mann Whitney U
-Test (also known as the Wilcoxon Rank Sum Test) to be sure of our results. Also,
-notice how some of these research questions call for a one-sided hypothesis test,
+Test (also known as the Wilcoxon Rank Sum Test) to be sure of our results.
+
+Also, notice how some of these research questions call for a one-sided hypothesis test,
 like the number of hits on ground balls, because we intuitively expect this number
@@
 
-Summarize the results.
+Insert description about Table~\ref{tab:table1} here.
+
+\begin{table}[tbp]
+  \caption{The results from the 9 tests}
+  \label{tab:table1}
+  \centering
+  \begin{tabular}[t]{lccccc}
+    \toprule
+    & 2023 & 2022 & \multicolumn{2}{c}{P-value}\\
+    \cmidrule(lr){4-5}
+    &          &                & \(t\)-test & Wilcoxon's\\
+    \midrule
+    Stolen Bases & 7.226 & 4.994 & 0.0000 & 0.0011 \\ 
+    Strikeouts & 85.445 & 81.731 & 0.1596 & 0.0000 \\ 
+    Walks & 32.876 & 30.279 & 0.0599 & 0.0791 \\ 
+    Batting Avg & 0.243 & 0.237 & 0.0274 & 0.0154 \\ 
+    Slugging \% & 0.401 & 0.384 & 0.0008 & 0.0007 \\ 
+    BABIP & 0.293 & 0.288 & 0.0793 & 0.0695 \\ 
+    xBA & 0.241 & 0.235 & 0.0006 & 0.0005 \\ 
+    Hits on GB & 27.345 & 26.399 & 0.3757 & 0.3212 \\ 
+    GIDP & 7.232 & 6.925 & 0.3523 & 0.5401 \\ 
+    \bottomrule
+  \end{tabular}
+\end{table}
 

### 2023-11-10 code/tables.R

@@ -1,0 +1,83 @@
+library(xtable)
+
+p_2023 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2023.csv")
+p_2022 <- read.csv("course-paper-or-presentation-lukenoel20/data/players_2022.csv")
+g_times <- read.csv("course-paper-or-presentation-lukenoel20/data/game_times.csv")
+
+# Performing a two-sample t-test and Wilcoxon Rank Sum test on each variable (game length is separate) and storing them for table later
+# Stolen Bases
+one_t <- t.test(p_2023$r_total_stolen_base, p_2022$r_total_stolen_base, alternative = 'greater')
+one_w <- wilcox.test(p_2023$r_total_stolen_base, p_2022$r_total_stolen_base, alternative = 'greater')
+r_one <- c(as.numeric(one_t$estimate[1]), as.numeric(one_t$estimate[2]), one_t$p.value, one_w$p.value)
+
+# Strikeouts
+two_t <- t.test(p_2023$strikeout, p_2022$strikeout)
+two_w <- wilcox.test(p_2023$strikeout, p_2022$strikout)
+r_two <- c(as.numeric(two_t$estimate[1]), as.numeric(two_t$estimate[2]), two_t$p.value, two_w$p.value)
+
+# Walks
+three_t <- t.test(p_2023$walk, p_2022$walk)
+three_w <- wilcox.test(p_2023$walk, p_2022$walk)
+r_three <- c(as.numeric(three_t$estimate[1]), as.numeric(three_t$estimate[2]), three_t$p.value, three_w$p.value)
+
+# Batting Average
+four_t <- t.test(p_2023$batting_avg, p_2022$batting_avg)
+four_w <- wilcox.test(p_2023$batting_avg, p_2022$batting_avg)
+r_four <- c(as.numeric(four_t$estimate[1]), as.numeric(four_t$estimate[2]), four_t$p.value, four_w$p.value)
+
+# Slugging Percentage
+five_t <- t.test(p_2023$slg_percent, p_2022$slg_percent)
+five_w <- wilcox.test(p_2023$slg_percent, p_2022$slg_percent)
+r_five <- c(as.numeric(five_t$estimate[1]), as.numeric(five_t$estimate[2]), five_t$p.value, five_w$p.value)
+
+# BABIP
+six_t <- t.test(p_2023$babip, p_2022$babip)
[…]

### 2023-11-13 code/tables.R
added code for game length table in tables.R file

@@ -80,4 +80,22 @@ print(table_object, include.rownames = TRUE)
 
 # NOW STARTING TABLE 2 (GAME LENGTH TABLE)
 
+# Convert AvgTime to numeric format (minutes)
+g_times$AvgTime_numeric <- as.numeric(sub(":.*", "", g_times$AvgTime)) + as.numeric(sub(".*:", "", g_times$AvgTime))/60
 
+# Dropping 2023 row from game length
+g_times <- g_times[-1, , drop = FALSE]
+
+# Perform  t-test and Wilcoxon on game length greater than 2023's which was 2.65 hours
+time_t <- t.test(g_times$AvgTime_numeric, mu = 2.65, alternative = "greater")
+time_w <- wilcox.test(g_times$AvgTime_numeric, mu = 2.65, alternative = "greater")
+
+# Data needed for latex table (will be inputted manually in tex file (easier))
+time2023 <- 2.65
+time2022_1998 <- mean(g_times$AvgTime_numeric)
+tpval_times <- time_t$p.value
+wpval_times <- time_w$p.value
+
+print(time2022_1998)
+print(tpval_times)
+print(wpval_times)

### 2023-11-13 manuscript/paper.tex
completed the Results section of manuscript

@@
 
-Insert description about Table~\ref{tab:table1} here.
+Table~\ref{tab:table1} summarizes the p-values for both the \(t\)-tests and Wilcoxon's
+tests for the research questions / hypotheses detailed in Section~\ref{sec:meth}.
+The difference in game length test was excluded from this table as this tested
+2023 versus 1998-2022 rather than just 2022, due to the data restrictions. Also,
+point estimates for each test were provided, one for the 2023 group and one for the
+2022 group.
+
+At the top of the table we see that the mean number of stolen bases for each player
+in 2023 is 7.226, and 4.994 in 2022 (these are the ``point estimates'') for the tests.
+This difference is statistically significant for both the \(t\)-test and Wilcoxon's test
+at $\alpha$ = 0.05, the significance level used for all of the tests in the table,
+as the p-values are 0.0000 (extremely small, rounded for presentation purposes) and
+0.0011 respectively. Adjusting the p-values to account for multiple tests using the
+Bonferroni correction also shows significance for both of these tests for stolen bases.
+This gives us ample evidence to conclude that the mean number of stolen bases in 2023
+significantly increased from 2022.
+
+On the other hand, the number of strikeouts did not seem to significantly change,
+as its \(t\)-test p-value is 0.1596. The Wilcoxon's test p-value is extremely small
+however, so the two tests are conflicting in their conclusions. Once we look at the
+95\% confidence interval (CI) of \((-1.46, 8.89)\) though, we see that it contains zero,
+meaning we cannot say the difference in the number of strikeouts year to year is significant.
+Moving down the table, other hypotheses that were not significant were Walks, BABIP,
+Hits on Ground Balls, and Ground into Double Play (GIDP) for both tests at the
+5\% significance level. Hits on Ground Balls was particularly surprising, as
+we expected that with no defensive shift anymore, more ground balls would become
+base hits.
+
+The difference in Batting Average, Slugging Percentage, and Expected Batting Average (xBA)
+from 2022 to 2023 were all statistically significant for both the \(t\)-test and
+Wilcoxon's Rank Sum test as their p-values were all less than $\alpha$ = 0.05. However,
+using the Bonferroni correction, the adjusted p-values for the Batting Average
+tests no longer were significant. So now there is only enough evidence to support
+that the difference from 2022 to 2023 in Slugging Percentage and xBA is significant.
+
+Also, for the game time data not shown in Table~\ref{tab:table1}, the average game
+length from 1998 to 2022 was 2.92 hours (2 hrs 55 mins), compared to only 2.65 hours
+(2 hrs 39 mins) in 2023. The \(t\)-test and Wilcoxon's test for the difference in
+these values yielded two extremely small p-values, so we have enough evidence to 
+conclude that the average game length is significantly shorter in 2023 compared to
+that in the previous 25 years. This was expected, as we assumed the pitch clock would
+speed up the game considerably.
 
 \begin{table}[tbp]
-  \caption{The results from the 9 tests}
+  \caption{The results from the 9 baseball statistics tests}
   \label{tab:table1}

### 2023-11-14 manuscript/paper.tex
finished Discussion section of manuscript

@@
 
-What are the main contributions again?
+This study provides valuable insights into the immediate effects of the 2023 MLB
+rule changes on various baseball statistics. Notably, there is robust evidence
+indicating a significant increase in the mean number of stolen bases during the
+2023 season. The introduction of a pitch clock and pitcher disengagement limits
+appears to have influenced baserunning strategies, fostering a more aggressive
+approach on the base paths. Additionally, offensive performance metrics such as
+slugging percentage and expected batting average demonstrated significant changes,
+suggesting that hitters adapted their approaches to capitalize on the new game dynamics.
 
-What are the limitations of this study?
+However, several limitations should be acknowledged. The study's focus on a single-season
+analysis, comparing 2023 to the previous season (2022), provides a snapshot of the
+immediate impacts but falls short of capturing potential trends that may have been
+happening over multiple seasons. External factors, such as individual player performance,
+team strategies, and other influences like basic season to season variation, could have
+contributed to observed differences, necessitating a more nuanced examination.
+Furthermore, data constraints, particularly the availability of detailed game-level
+data for certain metrics, limited the depth of the analysis.
 
-What are worth pursuing further in the future?
+In light of these limitations, future research directions are identified. Long-term
+analyses spanning multiple seasons would provide a more comprehensive understanding
+of the rule changes' lasting effects. Exploring player-level impacts and variations
+across teams could reveal nuanced insights into how individual athletes and teams
+adapted to the new rules. Additionally, a broader set of baseball metrics, including
+pitcher statistics like Earned Run Average (ERA), and defensive statistics like
+Outs Above Average (OAA), could contribute to a more holistic assessment of the rule
+changes' impact. Further research assessing fan engagement, satisfaction, and attendance
+would enhance our understanding of how these rule changes are perceived by the public.
+
+All in all, while this study sheds light on the immediate impacts, the complex and
+dynamic nature of baseball necessitates continuous monitoring and in-depth
+analysis to grasp the different effects of rule changes on the sport. The findings
+presented here serve as a great starting point for a more nuanced exploration of the
+evolving landscape of Major League Baseball.
 

### 2023-11-14 manuscript/paper.tex
changed the roadmap in the Intro section

@@
 The data used in the study will be presented in Section~\ref{sec:data}.
-The methods are described in Section~\ref{sec:meth}.
-The results are reported in Section~\ref{sec:resu}.
-A discussion concludes in Section~\ref{sec:disc}.
+The statistical methods and hypotheses are described in Section~\ref{sec:meth}.
+The results and findings are reported in Section~\ref{sec:resu}.
+An overall discussion concludes in Section~\ref{sec:disc}.
 

### 2023-11-14 manuscript/paper.tex
revised the abstract

@@
 their actual impact on various aspects of baseball performance and strategy is 
-not well understood. This study uses statistical methods to evaluate the effects
-of the rule changes on game duration, scoring, hitting, pitching, fielding, and
-baserunning. We compare the 2023 season data with the previous seasons using
-descriptive statistics, hypothesis testing, and regression analysis. INSERT
-major findings here.
+not well understood. By meticulously comparing the 2023 season data with preceding seasons, 
+utilizing descriptive statistics and hypothesis testing, the study unveils significant
+insights into the transformative effects of the 2023 MLB rule changes. The findings
+offer valuable perspectives for the future landscape of baseball.
 \end{abstract}
@@
 and know when to start stealing. We compare the 2023 season data with the previous
-seasons using descriptive statistics, hypothesis testing, and regression analysis.
+seasons using descriptive statistics and hypothesis testing.
 
